Large Linear Multi-output Gaussian Process Learning

### 2017-03-31 paper/paper.tex
minor paper updates

\begin{figure}[!ht]
\vskip 0.2in
\begin{center}
\centerline{\includegraphics[width=\columnwidth]{iterations.eps}}
\caption{Number of MVMs that MINRES must perform at each optimization iteration for a GP applied to the dataset in Section~\ref{fx2007-results}. The iteration cutoff is the number of training points $n$ in the dataset.
} % TODO no time right now but later could compare to LCG MVMs.
\label{fx2007-iterations}
\end{center}
\vskip -0.2in
\end{figure}

@@

In other words, all the work is done in the inversions $K_{|\bsth}^{-1}\tbr,K_{|\bsth}^{-1}\Ty$, which in turn rely on the MVM operation, discussed in Section~\ref{fast-mvm}. Since $K_{|\bsth}$ only enters computation as an MVM operator, the amount of memory consumed is dictated by its representation, which need not be dense. % this is a good transition paragraph, and motivates what follows well, but perhaps reworking the structure of the prior sections will make this (for lack of a better word) pipeline bottleneck mroe obvious? TODO any suggestions?

@@

% TODO get rid of all the linear terms below; delay big-O discussion until "asymptotics"?

@@

On a grid of inputs with $\TX=\TU$, the SKI interpolation vanishes with $W=I$. In this case, using \textsc{bt} alone leads to a faster algorithm --- applying the Chan block-Toeplitz preconditioner in a Krylov-subspace based routine has experimentally shown convergence using fewer MVMs \cite{chan1994circulant}.

@@

\begin{figure}[!ht]
\vskip 0.2in
\begin{center}
\centerline{\includegraphics[width=\columnwidth]{running_cutoff.eps}}
\caption{In green, we have 20\% of the rolling maximum norm. In red and blue are $\mcL$ (computed exactly and therefore unavailable during benchmarks) and $ \norm{\nabla\mcL}_\infty $, respectively. }
\label{fx2007-stop}
\end{center}
\vskip -0.2in
\end{figure}

@@

Here, experiments were carried out on ``random'' parameters with the same seed. % im not sure exactly what that means ("random parameters")? same seed? what is an experiment? might call it an application to simulated data? if we are desperate for space, make this a single sentence... instead of an itemized list?
In particular, we have the following parameters:
\begin{itemize}
\item $n$ total sample size across all outputs
\item $D$ number of outputs
\item $Q$ number of kernels $k_q$
\item $R$ average added rank
\item $\epsilon$ mean noise
\item \texttt{ktype} kernel type, one of \texttt{mat, periodic, rbf, mix} corresponding to Mat\'{e}rn-3/2, standard periodic\footnote{We define the periodic kernel as $k(r) = \exp \left(\frac{-\gamma}{2}\sin^2\frac{\pi r}{T}\right)$.}, and radial basis functions. \texttt{mix} refers to a mix of all three kernels.
Inverse length scales and periods for the kernels varied uniformly in log space from 1 to 10 with $Q$ samples.
\end{itemize}

Given this parameterization, we construct a random LMC kernel by sampling entries of each $A_q$ from a standard normal distribution, $\bsk_q$ from an inverse gamma with unit shape and scale, and independent noise $\bse$ for every output from an inverse gamma with unit scale and mean $\epsilon$. Inputs and outputs were independently generated from $\Unif[0,1]$ for benchmarking.

As expected from their asymptotic runtime, \textsc{sum}, \textsc{bt}, and \textsc{slfm} representations are complimentary in MVM speed for different configurations of $D,R,Q$ --- this results in sparse inversion computation that consistently outperforms Cholesky decomposition in runtime (Tab.~\ref{complement}).
For inverting systems, all computations were carried out until the residual $\ell_2$ norm was at most $10^{-4}$.

\begin{table}[!ht]
  \caption{The runtime in seconds for solving $K\Tx=\Ty$ for a random kernel $K$ constructed as in Section~\ref{empirical-rep} using MINRES for each of the kernel representations. For comparison, the \textsc{chol} representation is wallclock time to compute the Cholesky decomposition of the matrix, which must be constructed, and use this decomposition to invert the system. We averaged over five runs. In every run, we use $n=5,000$ simulated data points, $\texttt{mix}$ kernels, and $\epsilon=0.1$.
}
\label{complement}
\vskip 0.15in
\begin{center}
  \begin{small}
    \input{representation.tex}
\end{small}
\end{center}
\vskip -0.1in
\end{table}

@@

\begin{figure}[!ht]
\vskip 0.2in
\begin{center}
\centerline{\includegraphics[width=\columnwidth]{relgrad_l2.eps}}
\caption{Negative logarithm of the relative error in $\ell_2$ norm between exact and LLGP gradients. Higher is better, and extremal values are noted.  For each data point, the average was taken over five runs.}
\label{fig:relgrad}
\end{center}
\vskip -0.2in
\end{figure}

\begin{figure}[!ht]
\vskip 0.2in
\begin{center}
\centerline{\includegraphics[width=\columnwidth]{time_ratio.eps}}
\caption{Logarithm of the ratio of time required by the exact approach to that of LLGP for computing the gradient over all parameters from Eq.~\ref{llgradient}. Higher is better, and extremal values are noted. For each data point, the average was taken over five runs.}
\label{fig:gradtime}
\end{center}
\vskip -0.2in
\end{figure}

\begin{figure}[!ht]
\vskip 0.2in
\begin{center}
\centerline{\includegraphics[width=\columnwidth]{relalpha_l2.eps}}
\caption{Negative logarithm of the relative error in $\ell_2$ norm between exact and MINRES solutions to $K\Tx=\Ty$ for $\Tx$. For each data point, the average was taken over five runs.}
\label{fig:inverr}
\end{center}
\vskip -0.2in
\end{figure}

@@

We replicate the medium-sized dataset from COGP as an application to evaluate LLGP performance. The dataset consists of ten foreign currency exchange rates---CAD, EUR, JPY, GBP, CHF, AUD, HKD, NZD, KRW, and MXN---and three precious metals---XAU, XAG, and XPT---implying that $D=13$.\footnote{Data are from \texttt{http://fx.sauder.ubc.ca/data.html}} As in COGP, we retrieved the asset to USD rate, then used its reciprocal in all the results discussed below. The LLGP setting has $Q=1,R=2$, as recommended in \cite{alvarez2010efficient} for LMC models on this dataset; let this be the LMC model on LLGP. COGP roughly corresponds to the the SLFM model, which has a total of 94 hyperparameters, compared to 53 for LLGP. All kernels are RBF.

@@

\begin{table}[!ht]
  \caption{Average predictive performance and training time over $10$ runs for LLGP and COGP on the FX2007 dataset. Parenthesized values are standard error. LLGP was run with LMC set to $Q=1$, $R=2$, and $238$ interpolating points. COGP used a $Q=2$ kernel with $100$ inducing points}
\label{fx2007-tbl}
\vskip 0.15in
\begin{center}
  \begin{small}
    \begin{sc}
      \input{results_fx2007.tex}
\end{sc}
\end{small}
\end{center}
\vskip -0.1in
\end{table}

\begin{figure*}[!ht]
\vskip 0.2in
\begin{center}
\centerline{\includegraphics[width=\textwidth]{fx2007graph.pdf}}
\caption{Test outputs for the FX2007 dataset. COGP mean is black, with 95\% confidence intervals shaded in grey. LLGP mean is a solid red curve, with light green 95\% confidence intervals. Magenta points are in the training set, while blue ones are in the test set.} % TODO much better -- might also have 10 words on high level point: Predictions from LLGP on the test set are uniformly better across these three tasks. or something.
\label{fx2007-graph}
\end{center}
\vskip -0.2in
\end{figure*}

@@

Next, we replicate results from a weather dataset, a large time series used to validate COGP. In this dataset, $D=4$ weather sensors Bramblemet, Sotonmet, Cambermet, and Chimet record air temperature over five days in five minute intervals, with some dropped records due to equipment failure. Parts of Cambernet and Chimet are dropped for imputation, yielding $n=15,789$ training measurements and $374$ test measurements.

@@

\begin{figure}[!ht]
\vskip 0.2in
\begin{center}
\centerline{\includegraphics[width=\columnwidth]{m_time_nlpd.eps}}
\caption{Average and standard error of NLPD and runtime of the SLFM model on LLGP across over varying interpolating points. Every setting was run 10 times.}
\label{fig:llgpweather}
\end{center}
\vskip -0.2in
\end{figure}
\section{Conclusion}\label{conclusion}

>>>>

\begin{figure}[!ht]
\vskip 0.1in
\begin{center}
\centerline{\includegraphics[width=\columnwidth]{iterations.eps}}
\caption{Number of MVMs that MINRES must perform at each optimization iteration for a GP applied to the dataset in Section~\ref{fx2007-results}. The iteration cutoff is the number of training points $n$ in the dataset.
}
\label{fx2007-iterations}
\end{center}
\vskip -0.1in
\end{figure}

@@

In other words, all the work is done in the inversions $K_{|\bsth}^{-1}\tbr,K_{|\bsth}^{-1}\Ty$, which in turn rely on the MVM operation, discussed in Section~\ref{fast-mvm}. Since $K_{|\bsth}$ only enters computation as an MVM operator, the amount of memory consumed is dictated by its representation, which need not be dense.

@@

@@

On a grid of inputs with $\TX=\TU$, the SKI interpolation vanishes with $W=I$. In this case, using \textsc{bt} alone leads to a faster algorithm---applying the Chan block-Toeplitz preconditioner in a Krylov-subspace based routine has experimentally shown convergence using fewer MVMs \cite{chan1994circulant}.

@@

\begin{figure}[!ht]
\vskip 0.1in
\begin{center}
\centerline{\includegraphics[width=\columnwidth]{running_cutoff.eps}}
\caption{In green, we have 20\% of the rolling maximum norm. In red and blue are $\mcL$ (computed exactly and therefore unavailable during benchmarks) and $ \norm{\nabla\mcL}_\infty $, respectively. }
\label{fx2007-stop}
\end{center}
\vskip -0.1in
\end{figure}

@@

We evaluated the performance of the different kernel representations over various rank and parameterization settings. In particular, we have the following parameters: $n$ total sample size across all outputs, $D$ number of outputs, $Q$ number of kernels $k_q$, $R$ average added rank, $\epsilon$ mean noise, and \texttt{ktype} kernel type. Kernel type is one of \texttt{mat, periodic, rbf, mix} corresponding to Mat\'{e}rn-3/2, standard periodic\footnote{We define the periodic kernel as $k(r) = \exp \left(\frac{-\gamma}{2}\sin^2\frac{\pi r}{T}\right)$.}, and radial basis functions. \texttt{mix} refers to a mix of all three kernels.

Each kernel's inverse length scales and periods were selected by sampling uniformly in log space from 1 to 10 with $Q$ samples. Next, we construct a random LMC kernel by sampling entries of each $A_q$ from a standard normal distribution, $\bsk_q$ from an inverse gamma with unit shape and scale, and independent noise $\bse$ for every output from an inverse gamma with unit scale and mean $\epsilon$. Inputs and outputs were independently generated from $\Unif[0,1]$ for benchmarking.

As expected from their asymptotic runtime, \textsc{sum}, \textsc{bt}, and \textsc{slfm} representations are complimentary in MVM speed for different configurations of $D,R,Q$---this results in sparse inversion computation that consistently outperforms Cholesky decomposition in runtime (Tab.~\ref{complement}).
For inverting systems, all computations were carried out until the residual $\ell_2$ norm was at most $10^{-4}$.
\begin{table}[!ht]
  \caption{The runtime in seconds for solving $K\Tx=\Ty$ for a random kernel $K$ constructed as in Section~\ref{empirical-rep} using MINRES for each of the kernel representations. For comparison, the \textsc{chol} representation is wallclock time to compute the Cholesky decomposition of the matrix, which must be constructed, and use this decomposition to invert the system. We averaged over five runs. In every run, we use $n=5000$ simulated data points, $\texttt{mix}$ kernels, and $\epsilon=0.1$.
}
\label{complement}
\vskip 0.15in
\begin{center}
  \begin{small}
    \input{representation.tex}
\end{small}
\end{center}
\vskip -0.1in
\end{table}

@@

\begin{figure}[!ht]
\vskip 0.1in
\begin{center}
\centerline{\includegraphics[width=\columnwidth]{relgrad_l2.eps}}
\caption{Negative logarithm of the relative error in $\ell_2$ norm between exact and LLGP gradients. Higher is better, and extremal values are noted.  For each data point, the average was taken over five runs.}
\label{fig:relgrad}
\end{center}
\vskip -0.1in
\end{figure}

\begin{figure}[!ht]
\vskip 0.1in
\begin{center}
\centerline{\includegraphics[width=\columnwidth]{time_ratio.eps}}
\caption{Logarithm of the ratio of time required by the exact approach to that of LLGP for computing the gradient over all parameters from Eq.~\ref{llgradient}. Higher is better, and extremal values are noted. For each data point, the average was taken over five runs.}
\label{fig:gradtime}
\end{center}
\vskip -0.1in
\end{figure}

\begin{figure}[!ht]
\vskip 0.1in
\begin{center}
\centerline{\includegraphics[width=\columnwidth]{relalpha_l2.eps}}
\caption{Negative logarithm of the relative error in $\ell_2$ norm between exact and MINRES solutions to $K\Tx=\Ty$ for $\Tx$. For each data point, the average was taken over five runs.}
\label{fig:inverr}
\end{center}
\vskip -0.1in
\end{figure}

@@

We replicate the medium-sized dataset from COGP as an application to evaluate LLGP performance. The dataset consists of ten foreign currency exchange rates---CAD, EUR, JPY, GBP, CHF, AUD, HKD, NZD, KRW, and MXN---and three precious metals---XAU, XAG, and XPT---implying that $D=13$.\footnote{Data are from \url{http://fx.sauder.ubc.ca/data.html}} As in COGP, we retrieved the asset to USD rate, then used its reciprocal in all the results discussed below. The LLGP setting has $Q=1,R=2$, as recommended in \cite{alvarez2010efficient} for LMC models on this dataset; let this be the LMC model on LLGP. COGP roughly corresponds to the the SLFM model, which has a total of 94 hyperparameters, compared to 53 for LLGP. All kernels are RBF.

@@

\begin{table}[!ht]
  \caption{Average predictive performance and training time over $10$ runs for LLGP and COGP on the FX2007 dataset. Parenthesized values are standard error. LLGP was run with LMC set to $Q=1$, $R=2$, and $238$ interpolating points. COGP used a $Q=2$ kernel with $100$ inducing points.}
\label{fx2007-tbl}
\vskip 0.15in
\begin{center}
  \begin{small}
    \begin{sc}
      \input{results_fx2007.tex}
\end{sc}
\end{small}
\end{center}
\vskip -0.1in
\end{table}

\begin{figure*}[!ht]
\vskip 0.1in
\begin{center}
\centerline{\includegraphics[width=\textwidth]{fx2007graph.pdf}}
\caption{Test outputs for the FX2007 dataset. COGP mean is black, with 95\% confidence intervals shaded in grey. LLGP mean is a solid red curve, with light green 95\% confidence intervals. Magenta points are in the training set, while blue ones are in the test set. Notice LLGP variance corresponds to an appropriate level of uncertainty on the test set and certainty on the training set, as opposed to the uniform variance from COGP.}
\label{fx2007-graph}
\end{center}
\vskip -0.1in
\end{figure*}

@@

Next, we replicate results from a weather dataset, a large time series used to validate COGP. In this dataset, $D=4$ weather sensors Bramblemet, Sotonmet, Cambermet, and Chimet record air temperature over five days in five minute intervals, with some dropped records due to equipment failure. Parts of Cambernet and Chimet are dropped for imputation, yielding $n=15789$ training measurements and $374$ test measurements.

@@

\begin{figure}[!ht]
\vskip 0.1in
\begin{center}
\centerline{\includegraphics[width=\columnwidth]{m_time_nlpd.eps}}
\caption{Average and standard error of NLPD and runtime of the SLFM model on LLGP across over varying interpolating points. Every setting was run 10 times.}
\label{fig:llgpweather}
\end{center}
\vskip -0.1in
\end{figure}
\section{Conclusion}\label{conclusion}
